---
title: "EDUC 6003 Advanced Statistics"
author: "Hammed Akande"
format: html
editor: visual
---

# Chapter 1-

## Introduction to Reliability Index Analysis

This tutorial will serve as a Guide to Enhanced Reliability Estimation with R

### What is Reliability?

Reliability refers to the consistency and precision of measurement instruments. In psychological testing, it assesses the extent to which a test produces stable and consistent results.

#### Importance of Reliability

*Measurement Precision*: Ensures that the scores accurately reflect the true attributes being measured.

*Research Validity*: High reliability is a prerequisite for valid conclusions and replicable research findings.

*Error Reduction*: Minimizes the impact of random errors, enhancing the clarity of observed relationships between constructs.

#### Types of Reliability

*Internal Consistency*: Degree to which items within a test measure the same construct. 

*Test-Retest Reliability*: Consistency of scores over time. 

*Inter-Rater Reliability*: Agreement between different raters or observers.


## Classical Test Theory (CTT) and Coefficient Alpha

Classical Test Theory Overview: CTT posits that each observed score is composed of a true score and an error score:

$$ X = T + E $$

where:

𝑋= Observed score 𝑇= True score 𝐸= Error score

### Coefficient Alpha (Cronbach's Alpha)

This is a widely used measure of internal consistency reliability within CTT. It estimates the extent to which items on a test measure the same underlying construct.

$$
\alpha = \frac{N}{N-1} \left(1 - \frac{\sum \sigma^2_{E}}{\sigma^2_{X}}\right)
$$

Where:

$\alpha$ = Cronbach's Alpha 

𝑁 = Number of items in the test


$\sigma^2_{E}$ = Variance of the error scores


$\sigma^2_{X}$ = Variance of the observed total scores




#### Assumptions of Alpha

*Unidimensionality*: All items measure a single construct.

*Tau Equivalence*: Each item has the same true score variance.

*Independence of Errors*: Error terms are uncorrelated across items.


#### Limitations of Alpha

*Sensitivity to Tau Equivalence*: Violations can lead to underestimation or overestimation of reliability.

*Assumes Unidimensionality*: Not suitable for multidimensional scales without adjustments.

*Ignores Factor Structure*: Does not account for the underlying factor model of the test.


### Coefficient Omega

Coefficient Omega is a more robust alternative to Cronbach's Alpha, especially when the assumption of tau-equivalence (equal factor loadings) is violated. Omega takes into account the factor structure of the test, making it suitable for both unidimensional and multidimensional scales.




#### Types of Omega

*Omega Total (ωₜ*): Accounts for all common factors, both general and specific.

*Omega Hierarchical (ωₕ)*: Represents the proportion of variance attributable to a general factor alone.


Omega Total is given by:

$$
\omega_t = \frac{\sum_{i=1}^{k} \lambda_i}{\sum_{i=1}^{k} \lambda_i + \sum_{i=1}^{k} \theta_i}
$$


Where:

$\omega_t$  = Omega Total

$\lambda_i$  = Factor loading for item 

$\theta_i$  = Unique variance (error variance) for item 
   
$k$  = Total number of items


For a hierarchical model, Omega Hierarchical is:

$$
\omega_h = \frac{\lambda_g^2}{\lambda_g^2 + \sum_{i=1}^{k} \theta_i}
$$


where:


$\omega_h$  = Omega Hierarchical

$\lambda_g$  = Factor loading of the general factor

$\theta_i$  = Unique variance for item $i$


#### Advantages of Omega Over Alpha

*Factor Structure Incorporation*: Utilizes confirmatory factor analysis (CFA) to model the underlying structure.

*Less Sensitive to Tau Equivalence Violations*: Provides more accurate reliability estimates when tau equivalence is not met.

*Applicability to Multidimensional Scales*: Suitable for tests measuring multiple constructs.





## Calculations in R


Here, I will walk through the step-by-step process of calculating Cronbach's Alpha and Coefficient Omega using R. We'll use simulated data so it's easy to follow along (of course, you can apply these techniques to your own datasets).



Before we start, we need to install (and load) the necessary packages.

```{r}

# Install packages if not already installed (uncomment the code to run it)

#install.packages(c("psych", "lavaan", "semTools", "tidyverse"))

#Now, you can load the libraries
#library(psych)
#library(lavaan)
#library(semTools)
#library(tidyverse)

```

Alternatively, 

You could use the "pacman" package to easily install and load the packages above

```{r}

#pacman::p_load(psych,lavaan,semTools,tidyverse)

```



## Loading the (BFI) dataset

The Big Five Inventory (BFI) is a widely used dataset in personality psychology, measuring the five major dimensions of personality. We'll focus on the Openness trait for this tutorial.


```{r}

# Load the psych package
library(psych)

# Load the BFI dataset
data(bfi)

# load data file
bfi <- bfi
names(bfi)

# View the structure of the dataset
#str(bfi)




```


The dataset contains 2,800 participants with various demographic and personality-related variables. A1 to A5 correspond to items measuring the Openness trait.

### Selecting Openness Items

For this tutorial, we'll focus on five items measuring the Openness trait. These items are typically labeled as A1 to A5 in the BFI dataset.

#### a. Extracting Openness Items

```{r}

library(tidyverse)

# Select columns that contain "O" in their names and remove rows with NA values
oppeness <- bfi %>%
  select(matches("^O")) %>%
  drop_na()

head(oppeness)

```


Each A variable represents a Likert-type item measuring Openness. Responses range from 1 to 6, treated as continuous variables for analysis.


### Handling Missing Data

Before we continue, let's check any missing data.


```{r}

# Check for missing values
sum(is.na(oppeness))


```

In this dataset, there are no missing values in the Openness items. If missing values were present, then we need to address that.


We can remove the missing values like this. Note, since there's no missing values here, nothing to remove (but just letting you know).

```{r}

oppeness <- na.omit(oppeness)

any(is.na(oppeness))

```


## Calculating Cronbach's Alpha

Recall that Cronbach's Alpha (α) is a measure of internal consistency that measures how closely related a set of items are as a group.


### a. We can use the psych Package in R

```{r}

# Calculate Cronbach's Alpha using the psych package
#alpha_result <- psych::alpha(oppeness)

# Print the result
#print(alpha_result)



```


Because some items were negatively correlated with the first principal component, we probably 
should reverse them.  

```{r}

# Reversing item O2 and O5

q <- c("O2","O5")
reverse <- function(x){
  x.reversed <- 7 +0 - x
}

oppeness[, c("O2R", "O5R")] <- reverse(oppeness[,c("O2","O5")])

# compare original and reversed responses
oppeness[1:6,]

```

Now, we can compute the alpha

```{r}

# compute alpha coefficient 

alpha_result <- psych::alpha(oppeness[, -c(2,5)])

# Print the result
print(alpha_result)


```


Explanation:

raw_alpha: The Cronbach's Alpha coefficient.

std.alpha: Standardized alpha, similar to raw_alpha.

G6(smc): Generalizability theory estimate with squared multiple correlations.

average_r: Average inter-item correlation.

S/N: Signal-to-noise ratio.

ase: Asymptotic standard error.

Confidence Intervals: Lower and upper bounds for alpha.


As we can see, the Cronbach's Alpha for the Openness items is 0.6, indicating strong internal consistency. The average inter-item correlation is 0.24, moderate and within psychological scales.





## Confirmatory Factor Analysis (CFA)

Confirmatory Factor Analysis (CFA) is used to verify the factor structure of a set of observed variables. In our case, we assume a unidimensional structure.

### a. Specifying the CFA Model

Let's specify a one-factor model where all items load on a single latent factor.

```{r}

# Specify model
mod1f <- "
openness =~ O1 + O2R + O3 + O4 + O5R
"

```


### b. Fitting the CFA Model

Fit the Model Using lavaan:

```{r}
library(lavaan)
# Estimate model
fit.one.f <- cfa(mod1f, data=oppeness, std.lv=T, missing='direct', 
             estimator='MLR')

# The results can be viewed using the summary

summary(fit.one.f, fit.measures=T, standardized=T)



```




Explanation:

Estimator: Maximum Likelihood (ML). 

Test Statistics: The Chi-square test is significant (𝑝<0.001), indicating a poor fit.

Fit Indices:

CFI (0.94): above 0.90.

TLI (0.88): Below the threshold of 0.90.

RMSEA (0.068): Indicates relatively weak fit (≥0.05).

SRMR (0.029): Great fit (≤0.08).



```{r}

residuals(fit.one.f, type='cor')

```


The fit statistics reported under the Robust column in the summary indicate that this one-factor model has a marginal fit to the data; although the CFI of .94 suggests adequate fit, the TLI of .88 is lower than desired, and the RMSEA is somewhat high at 0.08.

These estimates of 𝜆̂ vary substantially (from .36 to.79), suggesting that tau equivalence is not tenable for this openness test and thus that ωu is a more appropriate reliability estimate than alpha.6 The ωu estimate can be obtained by executing the reliability function from the semTools package (i.e., semTools::reliability) on the fit1f one-factor model object:





```{r}

library(semTools)
reliability(fit.one.f)

```



The results listed in the omega and omega2 rows give the ωu estimate in which the denominator equals the model-implied variance of the total score X, and the results listed in the omega3 row are based on a variation in which the denominator equals the observed, sample variance of X, as described earlier.

The omega3 estimate is referred to as “hierarchical omega” in the help documentation for semTools::reliability (Jorgensen et al., 2020) and in Kelley and Pornprasertmanit (2016), whereas this Tutorial follows most of the psychometric literature (e.g., Rodriguez et al., 2016; Zinbarg et al., 2006) by reserving the term omega-hierarchical for estimates based on a bifactor model, which is presented later in this Tutorial.

Kelley and Pornprasertmanit (2016) reviewed different approaches to constructing CIs around omega estimates, ultimately recommending bootstrapping.






## Calculating Coefficient Omega

Recall that Coefficient Omega (ω) is a reliability estimate that accounts for the factor structure of the test, providing a more accurate measure than Cronbach's Alpha, especially in the presence of multidimensionality or when tau-equivalence is violated.


#Categorical omega

Because the factor loadings estimated from polychoric correlations represent the associations between the factor and latent-response variables rather than the observed item responses themselves, applying the ωu formula presented 
#earlier in this context would give the proportion of variance explained by the factor relative to the variance of the 
#sum of the latent-response variables instead of the total variance of the sum of the observed item responses; 
#that is, ωu would represent the reliability of a hypothetical, latent total score instead of the reliability of the 
#actual observed total score. To remedy this issue, Green and Yang (2009b) developed an alternative reliability estimate 
#that is rescaled into the observed total score metric. This reliability estimate for unidimensional categorical items 
#is referred to as ωu-cat.

#Yang and Green established that, compared with ωu, ωu-cat produces more accurate reliability estimates for total score 
#X with ordinal item-level data, especially when the univariate item-response distributions differ across items.

#Because the psychoticism items have a 4-point response scale, I fitted the one-factor model to the interitem 
#polychoric correlations using WLSMV, a robust weighted least squares estimator that is recommended over the ML 
#estimator for CFA with polychoric correlations (Finney & DiStefano, 2013).


```{r}
# read in data
library(rio)
potic <- import("potic.csv")

# Specify model
mod.one.fCat <- 'psyctcsm =~ DDP1 + DDP2 +
DDP3 + DDP4'


# Estimate model
fit.one.fCat <- cfa(mod.one.fCat, data=potic, std.lv=TRUE, ordered=T, estimator='WLSMV')

# Retrieving the results
summary(fit.one.fCat, fit.measures=TRUE, standardized=TRUE)


```





```{r}
#| echo: false
#2 * 2

#The `echo: false` option disables the printing of code (only output is displayed).

```

